Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints

### 2014-09-02 Compare_090214_statsmed_revisions.tex
final revisions

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

@@

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can be defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.

@@

When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased. However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities. Greater discussion of this point is given in Appendix A of the supplementary materials. When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued. When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.

@@

The $STG(t)$ is of interest for comparison of candidate surrogates regardless of the validity of assumption A6 as steeper quantile curves, and thus greater area between $R^t(v)$ and $(\rho_0(t) -\rho_1(t))$, correspond to better prediction of clinical treatment effect and a better candidate surrogate. Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=\max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$. This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting. Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-most left point on the $ROC^{t}(q)$ curve for the same time point $t$.

@@

As well, the same estimator we use at a fixed time point $t$ can be used to estimate the $STG(t)$ versus $t$ curve, for all $t$ greater than $\tau$ and less than the time of trial closure. This curve can be used to investigate time-variation in surrogate quality for a particular surrogate or for comparison. Curves, $STG(t)$ versus $t$, that have noticeable downward slopes may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining. To better understand the context of declining or increasing $STG(t)$ over time, the marginal risk differences over time should be graphed simultaneously. Figure \ref{STG} Panel A depicts a time-independent high quality surrogate and useless surrogate in a trial with declining marginal risk difference, Panel B depicts the same surrogates for a trial with increasing marginal risk difference.

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($\widetilde{STG}$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T. The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

@@

>>>>

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating a univariate biomarker as a PS of continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

@@

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can be defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well-defined  interpretations for our suggested summary statistics.

@@

When A6 is observably violated, the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased. However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities. Greater discussion of this point is given in Appendix A of the supplementary materials. When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued. When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.

@@

The $STG(t)$ is of interest for comparison of candidate surrogates regardless of the validity of assumption A6 as steeper quantile curves, and thus greater area between $R^t(v)$ and $(\rho_0(t) -\rho_1(t))$, correspond to better prediction of clinical treatment effect and a better candidate surrogate. Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=\max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$. This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting. Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-left-most point on the $ROC^{t}(q)$ curve for the same time point $t$.

@@

As well, the same estimator we use at a fixed time point $t$ can be used to estimate the $STG(t)$ versus $t$ curve, for all $t$ greater than $\tau$ and less than the time of trial closure. This curve can be used to investigate time-variation in surrogate quality for a particular surrogate or for comparison. Curves, $STG(t)$ versus $t$, that have noticeable downward trends may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining. To better understand the context of declining or increasing $STG(t)$ over time, the marginal risk differences over time should be graphed simultaneously. Figure \ref{STG} Panel A depicts a time-independent high quality surrogate and useless surrogate in a trial with declining marginal risk difference, Panel B depicts the same surrogates for a trial with increasing marginal risk difference.

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($\widetilde{STG}$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T. The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
By integrating over the Kaplan-Meier estimate, the $\widetilde{STG}$ statistic weights time points with observed events more heavily. This makes $\widetilde{STG}$ a sensible and empirically based summary of surrogate quality over all time-points of interest. The integrated $STG$ can be used to summarize the surrogate quality without specifying a fixed time point because it borrows information across times. It provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

@@

\ack The author are grateful to Dean Follmann and Ying Huang for their invaluable insights and helpful discussion. The authors also acknowledge both anonymous referees, whose comments helped us improve this article.
